Towards caring 15-minute neighbourhoods

### 2024-09-03 paper/Manuscript.qmd
Draft WGiT Presentation and edits to manuscript

<!--

•	Establishing definitive research questions and the work’s contributions. Us deciding on this together will help me sculpt from what’s there
  •	Thoughts on methods? Can we improve presentation?
  •	Thoughts on results? Can we improve presentation? What are insights that are significant to you? What do you want to know more about?
o	Thoughts on how I weighted care destinations
o	Thoughts on aggregation by ‘care category’ vs ‘specific type’ of destination
o	Thoughts on additional dimensions – i.e., the inclusion of results that only consider ‘grocery, errand, and health centric’ destinations (for people who have no dependents). For the time being, I left this consideration out and just demo _all_ destinations. Also, what about LICO-AT discussion?
  •	Critical thoughts on how to tie walking and cycling modes together. For now, I did not include cycling results, though I have the cycling travel time matrices. We should use the cycling results perhaps to demo how the range of care access can be extended for certain neighbourhoods. A potential ‘what-if’ scenario?
  •	Next steps

-->

@@

The 15-minute neighbourhood is a concept in planning that promotes urban spaces that are responsive to human needs and environmental sensibilities. These neighbuorhoods aim to accomplish this by ensuring that most daily necessities are not further away than a 15 minutes active trip [@Allam2022]. In practice, 15-minute neighborhoods are a form of normative accessibility [@paez_measuring_2012], and therefore related to accessibility analysis, an increasingly important tool that planners use to foster just and sustainable cities [@vale2023accessibility; @handyAccessibilityIdeaWhose2020]. Both the 15-minute neighborhood and accessibility measures are flexible enough to consider all destination types holistically, however, gendered examinations have been lacking in the literature. For instance, accessibility analyses have historically been employment-centric even when adopting a gendered perspective [@paez_jobs_2024].

To address this gender bias, this study investigates the 15-minute neighbourhood through the lens of Mobility of Care [@sanchezdemadariagaMobilityCareIntroducing2013]. This framework emphasizes the importance of travel related to unpaid work (care trips) in contrast to the better-studied travel to employment and leisure [@lawWomenTransportNew1999; @cresswell_gendered_2008; @Oestergaard1992; @levy1991towards; @levy2013travel]. While these trip types (work, care, and leisure) are essential, care trips are often relatively shorter-distance, proximate to residential/work/school, and comprise approximately 30% of adults’ daily trips [@sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory; @mejia2021mobility], thus fitting neatly within the concept of 15-minute neighbourhoods.

Our study casts the 15-minute neighbourhood onto the Mobility of Care framework. Empirically, we identify which areas in Hamilton, Canada, are 'caring 15-minute neighbourhoods'. To do so, a database of care destinations was compiled. In this database, care destinations include all places associated with sustaining household tasks needed for the reproduction of life namely: shopping (e.g., groceries), errands (e.g., libraries), health (e.g., dentist), and caring for dependents (e.g., schools). This database is used to estimate the number and mix of care-destinations that can be reached within a 15-minute walk-sheds from census centroids. We generate typologies to illustrate which neighbourhoods can and to what degree facilitate 15-minute access to care.

Our results indicate which 15-minute neighbourhoods are almost 'care-complete', that is, contain a sufficiently high mix of care destinations. Very few of these are found outside of the downtown core. Our study frames these neighbourhoods on a continuum of 'caring' and helps identify areas for policy intervention. Taken together, this study provides a theoretical bridge to connect and Mobility of Care, 15-minute neighbourhoods, and accessibility analysis. This should help to inform policy choices aimed at encouraging more just and sustainable mobility.

@@

Caring is one of the most fundamental activities in societies, since it consists of numerous activities that allow them to thrive. Caring activities are those that strive to meet the physical, psychological, and emotional needs of others [@ilo_care_2018]. Yet, it is one of the most unequal, undervalued, and even devalued activities worldwide. Traditionally, caring activities have been borne on women’s shoulders [@hayden_grand_2000; @hochschild_second_2012]. According to @ilo_care_2018, women and girls perform more than three-quarters of the total amount of unpaid care, a gender gap that varies geographically [@RN5].

This unequal share of caring responsibilities leads to multifaceted gendered differences: in career development, profession selection, contract type, pay gap, and time poverty, as recognized by various international organisations [@EIGE2016; @ilo_care_2018]. In fact, early feminists in the United States had already observed that domestic labor was the single most relevant cause of women’s inequality in society [@gordon_womans_1976; @hayden_grand_2000].

Care also has important implications for spatial and transportation planning. Indeed, researchers from various fields have highlighted that western modern industrial societies have imposed a physical separation of the private sphere from the public one, splitting production and reproduction, public and private spaces or simply work and home, concentrating on the economic outcome of being mobile [@cresswell_gendered_2008; @lawWomenTransportNew1999; @levy2013travel; @little_gender_1994; @tronto_toward_1990]. The man with legal rights, paid labour public sphere and woman in the private sphere, dependent, with family duties, personal relations and unpaid labour [@tronto_toward_1990]. In this line, @cresswell_gendered_2008 described how masculinity was related to mobile and active, whereas femininity to static. As an example of this dichotomy, transport planning historically framed non-work/school trips as 'non-mandatory' activities, downgrading these activities in comparison to work. In more recent years, planners have begun to resist this spatial separation of the public and private sphere to promote for sustainable cities. Proximity planning has become more normalized, at popular example of which is the 15-minute neighbourhood, in which residents can reach all necessary destinations within a 15-minute walk. The 15-minute neighbourhood concept, however, has yet to foreground care.

While much care labour takes place in the private and feminised sphere, completing care activities requires one to be mobile. In other words: caring for oneself and others requires trips to, amongst other places, grocery stores, day cares, elder care, and health care destinations.

From this research motivation, Sánchez de Madariaga coined the term ‘mobility of care’ in 2013 to refer to this travel or all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands [@sanchezdemadariagaMobilityCareIntroducing2013]. While decades of research had examined these unique household-serving trips, her seminal work was the first to consider all these trips as one category. In doing so, she showed how mobility of care is a significant proportion of daily travel. Indeed, in her original work, and in the studies that have followed, it appears that approximately 30% of adults’ trips are for care purposes [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. Perhaps unsurprisingly given the gendered division of care work discussed above, women have consistently been found to complete more mobility of care trips than men. In one study, mobility of care comprised 32% of women’s daily trips compared to 28% of men’s. While this gap was significant, it was found to be far greater in lower income households where women complete 20% more care trips than men [@ravensbergen2023exploratory]. Sánchez de Madariaga not only shows how important these mobility of care trips are, but also highlights the ways in which “mobility of care is systematically under-represented in any analysis of urban transport” (p. 37). Transport surveys and tools do not directly capture mobility of care, which re-enforces the idea that these trips are not a significant part of daily mobility.

@@

`r word_comment("TO ADD: some good feminist 15 minute cities. It's novel but movements exists in the Canadian context i.e., https://x.com/brenttoderian/status/1800371515877384527?s=48&t=QI0WSvV_FDb82ZGeLdJ9lQ. Also discuss Barcelona and Bogota (care blocks)", " ")`

Within the backdrop of urban planning and 'mobility of care', we connect our work to the 15-minute neighbourhood concept.

The spatial structure of cities and regions has been described in different ways and its advantages and disadvantages has been discussed with regard to a wide variety of issues, for example, regarding health, the environment, economic, social and equity impacts. After the traditional monocentric vision of the city with a core Centre Business District (CBD), other alternatives appeared, especially due to the expansion of cities caused by the automobile and motorized transport. The polycentric urban planning emerged based on the idea of sub-centers (or multiple business centers) with their own services and functions, well-connected among other surrounding areas, where the links between land-use and transport were optimised. Economic activities (such as retail or commercial office development) and job centers played an important role in shaping the spatial structure [@schneider_transit_1981; @sweet_are_2017]. This concept may have different interpretations depending on the spatial and geographical scales [@hall_polycentricity_2009].

@@

In the international context, there are different examples that highlight the importance of good access to different services and amenities within walkable distances. Probably one of the first cities to adopt the proximity approach was Portland. The Portland Plan [@city_of_portland_20minute_2010] was adopted by the City Council on April 25, 2012 aiming at fostering an inclusive city development based on prosperity, education, health, and equity for the following 30 years. The progress report states that equity should be achieved, an Office of Equity and Human Rights was established focusing on equitable service delivery to all residents. Nonetheless the progress report only measures progress on topics related to racial equity or people with disabilities [@portland_government_portland_2017]. `r word_comment("NOTE TO SELF: And so what? Is this a research gap?", " ")`

In Paris, the adoption of the 15-minute neighbourhood put into the spotlight this approach. They targeted six social functions that had to be easily accessible at any part of the city: housing, work, health care, groceries, education and leisure [@ville_de_paris_paris_2022]. In 2021, Ottawa also adopted a 15-Minutes approach in the new City of Ottawa Official Plan [@ville_dottawa_quartier_2021]. More recently, @moreno_introducing_2021 considers four dimensions on how the 15-minute neighbourhood framework: density (in terms of people per sq.km); diversity (including mixed-land use and diversity of people; the temporal and spatial proximity to essential services (residential areas, work, commercial areas, education centres, health facilities and other basic institutions) and digitalisation.  In this respect, the feminist perspective of the cities of proximity is still underestimated. Only few examples exist that specifically target this issue [@gil_sola_choose_2022; @macintyre_her_2022]. Given the impact of caring activities on everyday life, the chrono-urbanism should specifically target the analysis of caring activities.

@@

It appears that the car is the preferred travel mode for mobility of care. In one study, car use was higher for mobility of care trips than the commute to work (82.3% vs. 74.7%) [@ravensbergen2023exploratory]. Further, the rate of car use is higher amongst trips that are part of a chain than those that are not (85.28% vs. 78.09%) [@ravensbergen2023exploratory]. Some studies found that a car is viewed as the easier way to incorporate unpaid mobility of care into one’s daily travel as it provides convenience and mitigates concerns about traffic and stranger danger [@maciejewskaHaveChildrenThus2019, @carverParentalChauffeursWhat2013]. Others, however, have found that women are more likely to complete household-serving travel duties by alternative modes than men due to their lower car access [@RN24;@RN25]. It seems that car use may be preferred for mobility of care trips, but planning mobility of care for the car may leave those who do not have access to a car, many of them low-income women, with difficult mobility of care trips.

@@

Taken together, there appears to be a preference for travel by car for mobility of care trips. This is problematic, however, in at least two major ways. First, cars are less sustainable than other modes of transport. Normalizing car use for care trips does not advance the low-carbon transport transition that most cities are aiming for (at least in principle!). Since care trips tend to be short, they are also perfect candidate to be completed by bus, bicycle, or by foot. That is, if these alternative and sustainable modes are planned with care in mind. Second, there is equity concerns with assuming people are accessing care destinations by car. Many people, and low-income women in particular, do not have access to a car. Some are also experiencing a degree of spatial entrapment. Having safe and convenient access to care destinations using alternatives to the private car is especially important for those who must rely on these options regardless of whether they are well planned.

The objectives of the paper are as follows:

1)	City planning is moving to a people-centered approach focusing on accessibility to basic needs. However, chrono-urbanism stems from conventional methods that overlooks mobility of care perspectives. This paper presents a 15-minute cities methodological approach to assess mobility of care.
2)	We develop this approach using a mid-size typical Canadian city and a novel care destination dataset as case study
3)	Furthermore, the potential accessibility to and diversity of care destinations in the case study is analysed. We identify to what degree are neighbourhoods caring and/or complete. To do so, we make use of self organizing maps (SOM), an unsupervised machine learning data-dimensionality reduction technique to identify caring and complete neighbourhoods.
4)	We then apply the same SOM methodology to determine profiles of population groups to understand who enjoys these neighbourhoods and who does not.

# Data and Methods

@@

This study focuses on Hamilton, Ontario, a mid-sized city on the shore of Lake Ontario. Hamilton has a heterogenous land-use, with a populated and dense urban core, surrounded by suburban development, which is itself surrounded by rural communities made possible through a greenbelt. In this work, we analyse the residential parcel centroids, approximately ~140,000 locations. We aggregated the points at the level of Canadian Census Dissemination Area (DA) along with the population and population per parcel plots in @fig-Fig1.

@@

Hamilton also exhibits spatial disparities in social and economic indicators @fig-Fig2. The densely populated inner city is characterised by lower average incomes, and a higher prevalence of households living under the LICO. The suburban parts of the city tend to have a greater proportion of children and a lower proportion of one parent households.

@@

## Care destination dataset

A spatial dataset of care destinations for Hamilton was compiled; sources include provincial and municipal open data portals [@governmentofontarioOntarioDataCatalogue2023; @cityofhamiltonOpenData2023], Data Axle, a consumer dataset compiled of businesses and companies within Canada [@axledataConsumerData2023] and manual retrieved through Google Maps. Each destination was categorized based on the specific type of care being accessed following the mobility of care research by @sanchezdemadariagaMeasuringMobilitiesCare2019. Initial categories include child, elder,  grocery, health, and errand -centric destinations. For the purpose of this study, child- and elder- centric are combined to represent a 'dependent-centric' category. Category sources of data and preparation notes are detailed in @tbl-Tbl1. Their spatial distribution and sub-categories are visualised in @fig-Fig3.

>>>>

@@

The "15-Minute Neighbourhood" concept has been adopted by leaders around the world as a way to promote human-scale cities. The concept emphasizes a transport and land-use mix that supports short-trips to daily necessities by foot or bicycle from ones' residence. This study adopts the 15-minute neighbourhood concept through the lens of "Mobility of Care", a framework that foregrounds travel to care destinations, travel done predominately by women, in contrast to the better-studied travel to employment and leisure. While the 15-minute neighbourhood concept is flexible enough to consider all destination types, gendered examinations are lacking in the literature and no research to date has focused explicitly on care destinations. In this study, we address these gaps by identifying which areas in Hamilton, Canada are ‘caring 15-minute neighbourhoods'. To do so, a database of care destinations was compiled and used to estimate both the number (cumulative opportunities indicator) and diversity of mix (entropy measure) of destinations reachable within a 15-minute walk from all residential parcels. Parcel-level results were classified into 'caring' (quantity of reachable opportunities) and 'complete' (diversity of care destinations) neighbourhood typologies of varying degrees using the Self-Organizing Maps (SOM) machine learning technique. Inequalities in the relationship between the identified caring typologies and social-economic indicators were then explored using SOM. Our results suggest the majority of 'caring and complete' neighbourhoods are within the urban core where a higher proportion of lower-income and single-parent households reside. Further, certain neighbourhoods in suburban and rural communities are 'caring and somewhat complete', showing the potential to become 'complete' through possible urban policy intervention. Taken together, this study offers a theoretical bridge to connect Mobility of Care and the 15-minute neighbourhoods concept, a methodology to assess neighbourhoods on a 'caring and complete' continuum, and discussion of relevant gender mainstreaming policy interventions.

Keywords: Accessibility; 15-Minute Neighborhoods; Chrono-Urbanism; Cities of Proximity; Mobility of Care; Inequality; Gender Gap;

@@

The "15-Minute City" has been adopted by leaders as a way to promote human-scaled cities. As defined in @moreno_introducing_2021, the 15-Minute City is a urban planning model based in "chrono-urbanism", a theory that emphasizes the positive impact on quality of urban life when urban space becomes multi-rythmic [related to chronotopy @mulicekUrbanRhythmsChronotopic2015]. Multi-rhythmic urban space in opposition to industrial Fordism and single-use zoning [@mulicekUrbanRhythmsChronotopic2015; @morenoVilleQuartHeure2016]; the segregation of urban time and space for individual uses and mobility. The 15-Minute City is one that accommodates the ability to reach essential destinations such as shops, work, and healthcare within a walkable radius. It would allow individuals to reclaim time spent on car mobility and give way to sustainable modes; prompting urban spaces that are responsive to human needs and environmental sensibilities [@Allam2022]. The 15-Minute City generated much interest in planning practice and public discourse, especially after Paris mayor Anne Hidalgo proposed a 15-minute city plan in her 2020 re-election campaign. Though the term 15-Minute City is relatively new, the core of the concept is not.

Literature has drawn parallels from the 15-Minute City to Perry's debated Neighbourhood Unit Concept (NUC) from the 1930s [@kissfazekasCircleParadigms15minute2022]; a 'socio-spatial' normative scheme widely adopted by government agencies (and the real-estate community) in the Western world after the Second World War [@talenSocialSciencePlanned2017; @solow1969concept]. Pairing well with the objectives of planning agencies at the time, Perry's NUC would allow for efficient mass-building of cellular units that prioritized the perceived functional needs of women and children: each unit providing proximate access to an elementary school and sufficent supporting community facilities, shopping, parks and housing [@talenSocialSciencePlanned2017; @brodyNeighbourhoodUnitConcept2013]. The NUC primarily prioritized local service provision, though Perry had confidence in good design's contribution to 'neighbourhood spirit' [@hall2014cities]. From the NUC's conception until the end of the 1960s, planner's attempts to prescribe social meaning to the neighbourhood's physical form had been criticized to near extinction [@talenSocialSciencePlanned2017]. By social scientists, the overestimation on the impact of built environment on social life was critiqued. By other planners, the inability to agree on the specificity of the neighbourhood (i.e., population sized, and the type, quality, and quantity of amenity) as well as how neighbourhood units are connected to other neighbourhoods and the rest of the region. <!-- failing to combine work, home and cultre was a dilbirate outcome of state-sponsored capitalism X. As stated in [@talenSocialSciencePlanned2017],-->  In response to these criticisms, neighbourhood planning proponents have redefined deterministic terminology: prescribed physical forms is 'enabling' rather than engendering, and the provision of social contact 'opportunities' instead of determined social behaviour. The redefined bottom-up concept to community and neighbourhood planning has been adopted by New Urbanist from which the 15-Minute City stems. The 15-Minute City is created through bottom-up proximity-based processes of community planning [@kissfazekasCircleParadigms15minute2022].

<!-- These critiques were supported by case studies of planned neighbourhoods as... lead to social segregation of classes, destination functions, and automobile use. From it's onset, it was an efficient way to develop housing after the Second World War. The emphasis of proximity to some 'daily essentials' but disconnection from specialised activites (e.g., work) paved the way for car-based subrbia.  by capital of cities. Can the neighbourhood form might be seen as a vehicle for capitalist exploitation, gentrification and displacement??? The place of neighbourhood planning is actively being debated: can it's social crtiques be addressed in the new form as a "15-Minute City"? This study seeks to contribute to this decades old debate. -->

However, critiques of social scientists and planners from decades past still persist. Can physical form be planned to enable an improved quality of life? This line of reasoning has occupied urban planning research in recent years: what are the impacts of built environment on residence's outcomes. A popular tool to analyse this relationship has often been accessibility analysis: proximity to X amenity results in X outcomes for a neighboourhood. Accessibility is a way of quantifying the ease of opportunity interaction from some point in space, this indicator is becoming increasingly important for planners in the pursuit of just and sustainable city planning [@vale2023accessibility; @handyAccessibilityIdeaWhose2020]. As a amenity-provision concept of neighbourhood planning that seeks to enable an improvement on urban quality of life, the 15-Minute City lends well to analysis using accessibility tools.

Before discussing 'what outcome', we first ask 'what amenity' as the focus of this study. Often, accessibility analysis studies have focused on a single amenity type (X), often employment destinations and often focused on a scale beyond the neighbourhood. However, due to 15-Minute City's emphasis on proximate-level service provision, recent academic work on the topic has employed accessibility analysis to account for multiple 'local' amenities types, but seldom included employment. We argue that the inclusion of employment opportunities into the 15-Minute City conceptualisation is critical to it's definition. Why? We echo critics made by planners and social scientistics in decades past: neighbourhoods that did not mix work with service and culture within the neighbourhood, paved the way for car-based suburbia: cars allowed for the connection of neighbourhoods with their 'lower-order' amenities to areas with 'high-order' destinations like work (X). The lower- and higher- order framing of these destinations supported the continued segregation of urban life across many dimensions (X).

Combing all amenity types into the 15-Minute City concept is also a matter of gender equality. There is a branch of research that encompases this critique: Mobility of Care. This framework encourages the explicit consideration of destination types that serve unpaid labour (care destinations) into transport planning (as opposed to only considering destinations with paid work). Not considering these care destinations have travel time (and hence quality of life) implications on those who do care activites, commonly women and especially disadvantaged women. The consideration of _what_ amenity type is a gender issue.

From this position, this study aims to holistically investigate the ease of interacting with proximate amenities (accessibility) from the perspective of the 15-Minute City and the Mobility of Care framework. As mentioned, accessibility analysis have typically focused on employment (X) and recent 15-Minute City flavoured analysis focus on 'lower-order' amentities. We focus on a case study of the mid-sized City of Hamilton, Canada, and aim to classify areas by what degree they may be considered a '15-Minute Neighbourhood" based on the relative quantity of amenity access and diversity of access. We offer a novel methodology of holistically considering all proximate-amenities, offer empirical findings, and contribute theoretically to the ongoing debate on the definition of a 15-Minute City by foregrounding care destinations.

--------------

Caring activities meet the physical, psychological, and emotional needs of others, and are one of the most essential and fundamental activities in society [@ilo_care_2018]. Yet, they are the most unequal, undervalued, and even devalued activities worldwide. Traditionally, caring activities have been borne on women’s shoulders [@haydenGrandDomesticRevolution1982; @hochschild_second_2012]. According to @ilo_care_2018, women and girls perform more than three-quarters of the total amount of unpaid care, a gender gap that varies geographically [@RN5].

This unequal share of caring responsibilities leads to multifaceted gendered differences: in career development, profession selection, contract type, pay gap, and time poverty, as recognized by various international organisations [@EIGE2016; @ilo_care_2018]. In fact, early feminists in the United States had already observed that domestic labor was the single most relevant cause of women’s inequality in society [@gordon_womans_1976; @haydenGrandDomesticRevolution1982].

Care also has important implications for spatial and transportation planning. Indeed, researchers from various fields have highlighted that western modern industrial societies have imposed a physical separation of the private sphere from the public one, splitting production and reproduction, public and private spaces or simply work and home, concentrating on the economic outcome of being mobile [@cresswell_gendered_2008; @lawWomenTransportNew1999; @levy2013travel; @little_gender_1994; @tronto_toward_1990]. Men tend to be associated with legal rights, paid labour, and the public sphere while woman are associated with the private sphere, dependents, family duties, personal relations and unpaid labour [@tronto_toward_1990]. In this line, @cresswell_gendered_2008 described how masculinity has been constructed as mobile and active, whereas femininity is static. As an example of this dichotomy, transport planning has historically framed non-work/school trips as 'non-mandatory' activities, downgrading these activities in comparison to mandatory trips like work and school (XXX). In more recent years, planners have begun to resist the spatial separation of the public and private sphere to promote for sustainable cities. Proximity planning has become more normalized, a popular example is the 15-minute neighbourhood i.e., neighbourhoods where residents can reach all necessary destinations within a 15-minute active transport trip. The 15-minute neighbourhood concept, however, has yet to foreground care.

While much care labour takes place in the private and feminised sphere, completing care activities requires one to be mobile. In other words: caring for oneself and others requires trips to, amongst other places, grocery stores, day cares, elder care, and health care destinations. From this research motivation, Sánchez de Madariaga coined the term ‘mobility of care’ in 2013 to refer to all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands [@sanchezdemadariagaMobilityCareIntroducing2013]. While decades of research had examined these unique household-serving trips, her seminal work was the first to consider all these trips as one category. In doing so, she showed how mobility of care is a significant proportion of daily travel. Indeed, in her original work, and in the studies that have followed, it appears that approximately 30% of adults’ trips are for care purposes [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. Perhaps unsurprisingly given the gendered division of care work discussed above, women have consistently been found to complete more mobility of care trips than men. In one study, mobility of care comprised 32% of women’s daily trips compared to 28% of men’s. While this gap was significant, it was found to be far greater in lower income households where women complete 20% more care trips than men [@ravensbergen2023exploratory]. Sánchez de Madariaga not only shows how important these mobility of care trips are, but also highlights the ways in which “mobility of care is systematically under-represented in any analysis of urban transport” (p. 37). Transport surveys and tools do not directly capture mobility of care, which re-enforces the idea that these trips are not a significant part of daily mobility.

@@

`r word_comment("TO ADD: some good feminist 15 minute cities. It's novel but movements exists in the Canadian context i.e., https://x.com/brenttoderian/status/1800371515877384527?s=48&t=QI0WSvV_FDb82ZGeLdJ9lQ. Also discuss Barcelona and Bogota (care blocks) non academic sources: https://oecd-opsi.org/innovations/bogota-care-blocks/ AND https://www.npr.org/sections/goatsandsoda/2023/10/16/1205218707/how-bogota-cares-for-its-family-caregivers-from-dance-classes-to-job-training )", " ")`

<!-- The spatial structure of cities and regions has been described in different ways and its advantages and disadvantages has been discussed with regard to a wide variety of issues, for example, regarding health, the environment, economic, social and equity impacts. After the traditional monocentric vision of the city with a core Centre Business District (CBD), other alternatives appeared, especially due to the expansion of cities caused by the automobile and motorized transport. The polycentric urban planning emerged based on the idea of sub-centers (or multiple business centers) with their own services and functions, well-connected among other surrounding areas, where the links between land-use and transport were optimised. Economic activities (such as retail or commercial office development) and job centers played an important role in shaping the spatial structure [@schneider_transit_1981; @sweet_are_2017]. This concept may have different interpretations depending on the spatial and geographical scales [@hall_polycentricity_2009]. -->

@@

In the international context, there are different examples that highlight the importance of good access to different services and amenities within walkable distances. Probably one of the first cities to adopt the proximity approach was Portland. The Portland Plan [@city_of_portland_20minute_2010] was adopted by the City Council on April 25, 2012 aiming at fostering an inclusive city development based on prosperity, education, health, and equity for the following 30 years. The progress report describes the tasks of the established Office of Equity and Human Rights was established focusing on equitable service delivery to all residents. In the first example, equity is constrained to topics related to racial equity or people with disabilities [@portland_government_portland_2017]. `r word_comment("NOTE TO SELF: And so what? Is this a research gap?", " ")`

Later, other examples have highlighted the benefits of mobility of proximity without specifically targeting mobility of care, where a universal and neutral entity approach persists. In Paris, the adoption of the 15-minute neighbourhood put into the spotlight this approach. They targeted six social functions that had to be easily accessible at any part of the city: housing, work, health care, groceries, education and leisure [@ville_de_paris_paris_2022]. In 2021, Ottawa also adopted a 15-Minutes approach in the new City of Ottawa Official Plan [@ville_dottawa_quartier_2021]. More recently, @moreno_introducing_2021 considers four dimensions on how the 15-minute neighbourhood framework: density (in terms of people per sq.km); diversity (including mixed-land use and diversity of people; the temporal and spatial proximity to essential services (residential areas, work, commercial areas, education centres, health facilities and other basic institutions) and digitalisation.  In this respect, the feminist perspective of the cities of proximity is still underestimated. Only few examples exist that specifically target this issue [@gil_sola_choose_2022; @macintyre_her_2022]. Given the impact of caring activities on everyday life, the chrono-urbanism should specifically target the analysis of caring activities.

Within the backdrop of urban planning and ‘mobility of care’, we connect our work to the 15-minute neighbourhood concept.

Both the 15-minute neighborhood and accessibility measures are flexible enough to consider all destination types holistically, however, gendered examinations have been lacking in the literature. For instance, accessibility analyses have historically been employment-centric even when adopting a gendered perspective [@paez_jobs_2013]. Some 15-minute city analyses focus on care destinations, but they seem to not put care at the forefront, or to consider but one care destination (e.g., [@hosford15minuteCityReach2022] on grocery stores). Though few 15-minute neighbourhood analyses are explicitly feminist, their potential to build feminist communities have been noted outside of academia. For instance, Women in Urbanism have noted how women may benefit more than men from this 15-minute approach to planning. For instance, women tend to walk and use transit more than men, as well as to make more frequent, but shorter trips -and often as part of a chain (i.e., with stops along the way). These trip characteristics are all well served in 15-minute frameworks. 15-minute neighbourhoods also promote walkability, which results in more “eyes on the streets” and its associated safety, an important enabler of mobility for women and girls. Finally, because women tend to earn less money than men and cars are a large financial burden, planning paradigms that reduce car dependency, such as 15 minute cities, may on average benefit women more than men [@Bertolo2023].

@@

It appears that the car is the preferred travel mode for mobility of care. In one study, car use was higher for mobility of care trips than the commute to work (82.3% vs. 74.7%) [@ravensbergen2023exploratory]. Further, the rate of car use is higher amongst trips that are part of a chain than those that are not (85.28% vs. 78.09%) [@ravensbergen2023exploratory]. Some studies found that a car is viewed as the easier way to incorporate unpaid mobility of care into one’s daily travel as it provides convenience and mitigates concerns about traffic and stranger danger [@maciejewskaHaveChildrenThus2019, @carverParentalChauffeursWhat2013]. Others, however, have found that women are more likely to complete household-serving travel duties by alternative modes than men due to their lower car access [@RN24;@RN25]. It seems that car use may be preferred for mobility of care trips, but planning mobility of care for the car may leave those who do not have access to a car, many of them low-income women, with difficult mobility of care trips. Studies also observe disparities regarding gender mobility and type of activities, depending on the area of residence; women living in the city center, where more transport systems and activities are available, use more sustainable travel modes than the ones living in the outskirts, which tend to rely more on cars to carry out daily activities [@marquet2014walking; @mejia2021mobility; @montero2023role; @montero2023applying]. Outer areas are not as dense as in the downtown, and are basically car-oriented developments. Thus, studies observe that in the outer areas gender equity is confronted to sustainable mobility.

@@

Taken together, there appears to be a preference for travel by car for mobility of care trips in many geographical contexts. This is problematic, however, in at least two major ways. First, cars are less sustainable than other modes of transport. Normalizing car use for care trips does not advance the low-carbon transport transition that most cities are aiming for (at least in principle!). Since care trips tend to be short, they are also perfect candidate to be completed by bus, bicycle, or by foot. That is, if these alternative and sustainable modes are planned with care in mind. Second, there is equity concerns with assuming people are accessing care destinations by car. Many people, and low-income women in particular, do not have access to a car. Some are also experiencing a degree of spatial entrapment. Having safe and convenient access to care destinations using alternatives to the private car is especially important for those who must rely on these options regardless of whether they are well planned.

Taken together, city planning is moving to a people-centered approach focusing on accessibility to basic needs. However, chrono-urbanism stems from conventional methods that overlooks mobility of care perspectives. To counter this, this paper presents a 15-minute cities methodological approach to assess mobility of care. We develop this approach using a mid-size typical Canadian city and a novel care destination dataset as case study. The objectives of the paper are as follows:

1)	To examine the potential accessibility to, and diversity of, care destinations in the case study. We identify to what degree are neighbourhoods caring and/or complete. To do so, we make use of self organizing maps (SOM), an unsupervised machine learning data-dimensionality reduction technique to identify caring and complete neighbourhoods.
2)	We then apply the same SOM methodology to determine profiles of population groups to understand who enjoys these neighbourhoods and who does not.

# Data

@@

This study focuses on Hamilton, Ontario, a mid-sized city on the shore of Lake Ontario. Hamilton has a heterogenous land-use, with a populated and dense urban core, surrounded by suburban development, which is itself surrounded by rural communities. The Niagara Escarpment runs through Hamilton, and results in a city with two key elevations: the dense lower city that contains the downtown core and the elevated suburban development referred to as ‘the Mountain’ (ADD THIS TO FIGURES). In this work, we analyse the residential parcel centroids, approximately ~140,000 locations. We aggregated the points at the level of Canadian Census Dissemination Area (DA) along with the population and population per parcel plots in @fig-Fig1.

@@

Hamilton also exhibits spatial disparities in social and economic indicators @fig-Fig2. The densely populated inner city is characterised by lower average incomes, and a higher prevalence of households living under the low-income cuttoff thresholds (LICO). The suburban parts of the city tend to have a greater proportion of children and a lower proportion of one parent households.

@@

## Care destination dataset
A spatial dataset of care destinations for Hamilton was compiled; sources include provincial and municipal open data portals [@governmentofontarioOntarioDataCatalogue2023; @cityofhamiltonOpenData2023], Data Axle, a consumer dataset compiled of businesses and companies within Canada [@axledataConsumerData2023] and manual retrieved through Google Maps. Each destination was categorized based on the specific type of care being accessed following the mobility of care research by @sanchezdemadariagaMeasuringMobilitiesCare2019. Initial categories include child, elder,  grocery, health, and errand -centric destinations. For the purpose of this study, child- and elder- centric are combined to represent a 'dependent-centric' category. Category sources of data and preparation notes are detailed in @tbl-Tbl1. Their spatial distribution and sub-categories are visualised in @fig-Fig3.

```{r creating-table-content}
dat <- data.frame(matrix(c("Depedent-centric",
                           "[@opendatahamiltonEducationalInstitutions2022; @governmentofontarioOntarioDataCatalogue2023; @cityofhamiltonChildCareRegistry2023; @opendatahamiltonParksDatabase2022; @opendatahamiltonRecreationCommunityCentres2022; @ontarioministryofhealthgeohubMinistryHealthService2023]",
                           "Schools, daycares, and community centres, recreation centres, parks, senior centres, long-term care homes, and retirement homes: 1,265 locations are included. After manual review, all locations that typically do not serve children were removed including: Post-Secondary, Adult-Learning Centres, Group Homes, and Foster Care Centres. Further, through examination some Section 23 institutions defined as _“centres for children who cannot attend school to meet the needs of care or treatment, and rehabilitation”_ [@governmentofontarioministryofeducationFundingEducationCommunity2023], were kept due to their innate connection to care.",
                           "Grocery-centric",
                           "[@axledataConsumerData2023]",
                           "Grocery stores, namely a place a household could buy groceries ranging from convenience stores to large retail stores: 381 destinations are identified. Data is filtered by Company Name, Suite Number, Address, City, Province, Phone Number and Postal Code. The type was then identified e.g., grocers specialty foods, grocers retail, grocer health food, grocer wholesale, grocer curbside, grocer delicatessen wholesale, grocer convenience. Data was cross-referenced to ensure all included locations were operational and legitimate grocery stores.",
                           "Health-centric",
                           "[@ontarioministryofhealthgeohubMinistryHealthService2023; @hamiltonniagarahaldimandbranthealthlineWalkInMedicalClinics2023]",
                           "Hospitals, pharmacies, clinics, and dentist offices: 421 destinations are identified. Hospitals and pharmacies were retrieved while clinics and dentistry clinics were manually scraped from a healthcare services database and checked via Google Maps to remove non-operational locations and confirm dentistry-orientation. ",
                           "Errand-centric",
                           "Hamilton libraries [@opendatahamiltonLibraries2022], post office locations [@axledataConsumerData2023;@canadapostPostOfficeLocator2023], and datasets of all national bank chains [@bankofmontrealBankMontrealBranch2023; @thehongkongandshanghaibankingcorporationlimitedhsbcHSBCBranchABM2023; @nationalbankNationalBankBranches2023; @royalbankofcanadaRBCBranchABM2023; @scotiabankScotiabankABMBranch2023; @thetorontodominionbankTDBankBranch2023].",
                                                      "Libraries, post offices, and banks: 158 destinations are identified. Post offices are retrieved from a mix of databases, and duplicates are removed. Banks are also derived from Data Axle and then cross-referenced to ensure data quality with a Bank Locator website for all national banking firms."
                           ), ncol = 3, byrow = T))
```
